---
title: 'Chapter 6 in ISLR: Part 2 Regularization'
author: "Laura Kapitula"
date: "11/23/2021"
output:
    html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
```


This material on  Ridge Regression and the Lasso in R comes from Chapter 6 of "Introduction to Statistical Learning with Applications in R" by Gareth James, Daniela Witten, Trevor Hastie and Robert Tibshirani. I used code that did the lab  in `tidyverse` format by Amelia McNamara and R. Jordan Crouser at Smith College. from 
<http://www.science.smith.edu/~jcrouser/SDS293/labs/lab10.Rmd > to implement the example in R.

Note, you can try the tidymodels way at, https://emilhvitfeldt.github.io/ISLR-tidymodels-labs/linear-model-selection-and-regularization.html
but I was having some issues getting the plots and could not figure out how to fix it.

## Ridge regression and Lasso

* The subset selection methods use least squares to fit a
linear model that contains a subset of the predictors.

*As an alternative, we can fit a model containing all $p$
predictors using a technique that constrains or regularizes
the coefficient estimates, or equivalently, that shrinks the
coefficient estimates towards zero.

* It may not be immediately obvious why such a constraint
should improve the fit, but it turns out that shrinking the
coefficient estimates can significantly reduce their variance.

## Ridge Regression 
Recall that the least squares fitting procedure estimates
$\beta_0, \beta_1, ..., \beta_p$ using values that minimize

$$\sum_{i=1}^{n} \left( y_i - \beta_0 - \sum_{j=1}^{p} \beta_j x_{ij}    \right) ^ 2  $$

 In contrast, the ridge regression coefficient estimates $\hat{\beta}^R$ are the values that minimize:

$$\sum_{i=1}^{n} \left( y_i - \beta_0 - \sum_{j=1}^{p} \beta_j x_{ij}    \right) ^ 2 + \lambda \sum_{j=1}^{p} \beta_j^2 $$
where $\lambda \ge 0 $ is a tuning parameter, to be determined
separately.

* As with least squares, ridge regression seeks coefficient
estimates that fit the data well, by making the RSS small.

* However, the second term, 

$$ \lambda \sum_{j=1}^{p} \beta_j^2 $$

 is called a shrinkage penalty, is small when $\beta_1, \beta_1, ..., \beta_p$ are close to zero, and so it
has the effect of shrinking the estimates of $\beta_j$ towards zero.

* The tuning parameter $\lambda$ serves to control the relative impact of these two terms on the regression coefficient
estimates.

* Selecting a good value for $\lambda$ is critical and cross-validation is used for this.


```{r, results=FALSE, warning=FALSE, message=FALSE}
library(ISLR)
library(glmnet)
library(dplyr)
library(tidyr)
```

We will use the `glmnet` package in order to perform ridge regression and
the lasso. The main function in this package is `glmnet()`, which can be used
to fit ridge regression models, lasso models, and more. This function has
slightly different syntax from other model-fitting functions that we have
encountered thus far in this book. In particular, we must pass in an $x$
matrix as well as a $y$ vector, and we do not use the $y \sim x$ syntax.


```{r}
Hitters = na.omit(Hitters)
```

We will now perform ridge regression and the lasso in order to predict `Salary` on
the `Hitters` data. Let's set up our data:


```{r}
x = model.matrix(Salary~., Hitters)[,-1] # trim off the first column
                                         # leaving only the predictors
y = Hitters %>%
  select(Salary) %>%
  unlist() %>%
  as.numeric()
```

The `model.matrix()` function is particularly useful for creating $x$; not only
does it produce a matrix corresponding to the 19 predictors but it also
automatically transforms any qualitative variables into dummy variables.
The latter property is important because `glmnet()` can only take numerical,
quantitative inputs.

The `glmnet()` function has an alpha argument that determines what type
of model is fit. If `alpha = 0` then a ridge regression model is fit, and if `alpha = 1`
then a lasso model is fit. We first fit a ridge regression model:


```{r}
grid = 10^seq(10, -2, length = 100)
ridge_mod = glmnet(x, y, alpha = 0, lambda=grid)
```

By default the `glmnet()` function performs ridge regression for an automatically
selected range of $\lambda$ values. However, here we have chosen to implement
the function over a grid of values ranging from $\lambda = 10^{10}$ to $\lambda = 10^{-2}$, essentially covering the full range of scenarios from the null model containing
only the intercept, to the least squares fit. 

As we will see, we can also compute
model fits for a particular value of $\lambda$ that is not one of the original
grid values. Note that by default, the `glmnet()` function standardizes the
variables so that they are on the same scale. To turn off this default setting,
use the argument `standardize = FALSE`.

### Ridge regression: scaling of predictors
* The standard least squares coefficient estimates are scale
equivariant.

* In contrast, the ridge regression coefficient estimates can
change substantially when multiplying a given predictor by
a constant, due to the sum of squared coefficients term in
the penalty part of the ridge regression objective function.

* Therefore, it is best to apply ridge regression after
standardizing the predictors.

Associated with each value of $\lambda$ is a vector of ridge regression coefficients,
stored in a matrix that can be accessed by `coef()`. In this case, it is a $20 \times 100$
matrix, with 20 rows (one for each predictor, plus an intercept) and 100
columns (one for each value of $\lambda$).


```{r}
plot(ridge_mod)
dim(coef(ridge_mod))
 # Draw plot of coefficients
```


```{r}
plot(ridge_mod, xvar = "lambda", label = TRUE)
```
We expect the coefficient estimates to be much smaller, in terms of $l_2$ norm,
when a large value of $\lambda$ is used, as compared to when a small value of $\lambda$ is
used. These are the coefficients when $\lambda = 11498$, along with their $l_2$ norm:


```{r}
ridge_mod$lambda[50] #Display 50th lambda value
coef(ridge_mod)[,50] # Display coefficients associated with 50th lambda value
sqrt(sum(coef(ridge_mod)[-1,50]^2)) # Calculate l2 norm
```

In contrast, here are the coefficients when $\lambda = 705$, along with their $l_2$
norm. Note the much larger $l_2$ norm of the coefficients associated with this
smaller value of $\lambda$.


```{r}
ridge_mod$lambda[60] #Display 60th lambda value
coef(ridge_mod)[,60] # Display coefficients associated with 60th lambda value
sqrt(sum(coef(ridge_mod)[-1,60]^2)) # Calculate l2 norm
```

We can use the `predict()` function for a number of purposes. For instance,
we can obtain the ridge regression coefficients for a new value of $\lambda$, say 50:


```{r}
predict(ridge_mod, s = 50, type = "coefficients")[1:20,]
```

We now split the samples into a training set and a test set in order
to estimate the test error of ridge regression and the lasso.

```{r}
set.seed(1)

train = Hitters %>%
  sample_frac(0.5)

test = Hitters %>%
  setdiff(train)

x_train = model.matrix(Salary~., train)[,-1]
x_test = model.matrix(Salary~., test)[,-1]

y_train = train %>%
  select(Salary) %>%
  unlist() %>%
  as.numeric()

y_test = test %>%
  select(Salary) %>%
  unlist() %>%
  as.numeric()
```

Next we fit a ridge regression model on the training set, and evaluate
its MSE on the test set, using $\lambda = 4$. Note the use of the `predict()`
function again: this time we get predictions for a test set, by replacing
`type="coefficients"` with the `newx` argument.


```{r}
ridge_mod = glmnet(x_train, y_train, alpha=0, lambda = grid, thresh = 1e-12)
ridge_pred = predict(ridge_mod, s = 4, newx = x_test)
mean((ridge_pred - y_test)^2)
```

The test MSE is 101242.7. Note that if we had instead simply fit a model
with just an intercept, we would have predicted each test observation using
the mean of the training observations. In that case, we could compute the
test set MSE like this:


```{r}
mean((mean(y_train) - y_test)^2)
```

We could also get the same result by fitting a ridge regression model with
a very large value of $\lambda$. Note that `1e10` means $10^{10}$.


```{r}
ridge_pred = predict(ridge_mod, s = 1e10, newx = x_test)
mean((ridge_pred - y_test)^2)
```

So fitting a ridge regression model with $\lambda = 4$ leads to a much lower test
MSE than fitting a model with just an intercept. We now check whether
there is any benefit to performing ridge regression with $\lambda = 4$ instead of
just performing least squares regression. Recall that least squares is simply
ridge regression with $\lambda = 0$.

\* Note: In order for `glmnet()` to yield the **exact** least squares coefficients when $\lambda = 0$,
we use the argument `exact=T` when calling the `predict()` function. Otherwise, the
`predict()` function will interpolate over the grid of $\lambda$ values used in fitting the
`glmnet()` model, yielding approximate results. Even when we use `exact = T`, there remains
a slight discrepancy in the third decimal place between the output of `glmnet()` when
$\lambda = 0$ and the output of `lm()`; this is due to numerical approximation on the part of
`glmnet()`.


```{r}
ridge_pred = predict(ridge_mod, s = 0, newx = x_test)
mean((ridge_pred - y_test)^2)

lm(Salary~., data = train)
predict(ridge_mod, s = 0,  type="coefficients")[1:20,]
```

It looks like we are indeed improving over regular least-squares! Side note: in general, if we want to fit a (unpenalized) least squares model, then
we should use the `lm()` function, since that function provides more useful
outputs, such as standard errors and $p$-values for the coefficients.

Instead of arbitrarily choosing $\lambda = 4$, it would be better to
use cross-validation to choose the tuning parameter $\lambda$. We can do this using
the built-in cross-validation function, `cv.glmnet()`. By default, the function
performs 10-fold cross-validation, though this can be changed using the
argument `folds`. Note that we set a random seed first so our results will be
reproducible, since the choice of the cross-validation folds is random.


```{r}
set.seed(1)
cv.out = cv.glmnet(x_train, y_train, alpha = 0) # Fit ridge regression model on training data
bestlam = cv.out$lambda.min  # Select lamda that minimizes training MSE
bestlam
```

Therefore, we see that the value of $\lambda$ that results in the smallest cross-validation
error is 339.1845 We can also plot the MSE as a function of $\lambda$:


```{r}
plot(cv.out) # Draw plot of training MSE as a function of lambda
```

What is the test MSE associated with this value of
$\lambda$?


```{r}
ridge_pred = predict(ridge_mod, s = bestlam, newx = x_test) # Use best lambda to predict test data
mean((ridge_pred - y_test)^2) # Calculate test MSE
```

This represents a further improvement over the test MSE that we got using
$\lambda = 4$. Finally, we refit our ridge regression model on the full data set,
using the value of $\lambda$ chosen by cross-validation, and examine the coefficient
estimates.


```{r}
out = glmnet(x, y, alpha = 0) # Fit ridge regression model on full dataset
predict(out, type = "coefficients", s = bestlam)[1:20,] # Display coefficients using lambda chosen by CV
```

As expected, none of the coefficients are exactly zero - ridge regression does not
perform variable selection!

## The Lasso

Ridge regression does have one obvious disadvantage:
unlike subset selection, which will generally select models
that involve just a subset of the variables, ridge regression
will include all p predictors in the final model

* The Lasso is a relatively recent alternative to ridge
regression that overcomes this disadvantage. The lasso
coefficients,  estimates $\hat{\beta}^L$ are the values that minimize:

$$\sum_{i=1}^{n} \left( y_i - \beta_0 - \sum_{j=1}^{p} \beta_j x_{ij}    \right) ^ 2 + \lambda \sum_{j=1}^{p} | \beta_j |$$
where $\lambda \ge 0 $ is a tuning parameter, to be determined
separately.

* As with least squares, Lasso seeks coefficient
estimates that fit the data well, by making the RSS small.

* The second term, 

$$ \lambda \sum_{j=1}^{p} | \beta_j|$$

 is a shrinkage penalty, based on the $l_1$ norm.  When we add up the squared terms it is a $l_2$ norm .  For the Lasso, the  $\beta_1, \beta_1, ..., \beta_p$  can shrink the estimates of $\beta_j$ not just towards zero, but make them exactly equal to zero.

* However, in the case of the lasso, the penalty has the effect of forcing some of the coefficient estimates to be
exactly equal to zero when the tuning parameter $\lambda$ is sufficiently large.

* Hence, much like best subset selection, the lasso performs variable selection.
* We say that the lasso yields sparse models (does not use all predictors).

* Selecting a good value for $\lambda$ is critical and cross-validation is used for this.


We saw that ridge regression with a wise choice of $\lambda$ can outperform least
squares as well as the null model on the Hitters data set. We now ask
whether the lasso can yield either a more accurate or a more interpretable
model than ridge regression. In order to fit a lasso model, we once again
use the `glmnet()` function; however, this time we use the argument `alpha=1`.
Other than that change, we proceed just as we did in fitting a ridge model:

```{r}
lasso_mod = glmnet(x_train, 
                   y_train, 
                   alpha = 1, 
                   lambda = grid) # Fit lasso model on training data

plot(lasso_mod)    # Draw plot of coefficients
```

Notice that in the coefficient plot that depending on the choice of tuning
parameter, some of the coefficients are exactly equal to zero. We now
perform cross-validation and compute the associated test error:


```{r}
set.seed(1)
cv.out = cv.glmnet(x_train, y_train, alpha = 1) # Fit lasso model on training data
plot(cv.out) # Draw plot of training MSE as a function of lambda
bestlam = cv.out$lambda.min # Select lamda that minimizes training MSE
lasso_pred = predict(lasso_mod, s = bestlam, newx = x_test) # Use best lambda to predict test data
mean((lasso_pred - y_test)^2) # Calculate test MSE
```

This is substantially lower than the test set MSE of the null model and of
least squares, and very similar to the test MSE of ridge regression with $\lambda$
chosen by cross-validation.

However, the lasso has a substantial advantage over ridge regression in
that the resulting coefficient estimates are sparse. Here we see that 12 of
the 19 coefficient estimates are exactly zero:


```{r}
out = glmnet(x, y, alpha = 1, lambda = grid) # Fit lasso model on full dataset
lasso_coef = predict(out, type = "coefficients", s = bestlam)[1:20,] # Display coefficients using lambda chosen by CV
lasso_coef
```

Selecting only the predictors with non-zero coefficients, we see that the lasso model with $\lambda$
chosen by cross-validation contains only seven variables:


```{r}
lasso_coef[lasso_coef != 0] # Display only non-zero coefficients
```