---
title: "Regression and Other Stories: Residuals"
author: "Laura Kapitula"
date: "`r format(Sys.Date())`"
output:
  html_document:
    theme: readable
    toc: true
    toc_depth: 2
    toc_float: true
    code_download: true
---
Plotting the data and fitted model. See Chapter 11 in Regression
and Other Stories.
 Edited version of original file by: Andrew Gelman, Jennifer Hill, Aki Vehtari 

-------------


```{r setup, include=FALSE}
knitr::opts_chunk$set(message=FALSE, error=FALSE, warning=FALSE, comment=NA)
# switch this to TRUE to save figures in separate files
savefigs <- FALSE
```

#### Load packages

```{r }
library("rstanarm")
library(tidyverse)
```

## Simple model with const term, one pre-treatment predictor, and treatment indicator

#### Fake data

```{r }
N <- 100
x <- runif(N, 0, 1)
z <- sample(c(0, 1), N, replace=TRUE)
a <- 1
b <- 2
theta <- 5
sigma <- 2
y <- a + b*x + theta*z +  rnorm(N, 0, sigma)
fake <- data.frame(x=x, y=y, z=z)
summary(fake)
```


```{r}

ggplot(fake, aes(x=x, y=y, shape=as.factor(z), color=as.factor(z)) )+
  geom_point() +
    labs(x = "Pre-treatment predictor, x", y = "Outcome, y")
```

#### Model

```{r }
fit <- stan_glm(y ~ x + z, data=fake, refresh = 0)
print(fit)
```

#### Add predicted lines to our plot.
```{r }

#predicted is the y-hats, predicted values, using the model
#then we calculate the residuals as the actual value - the predicted value. 
fake2 <- fake %>%
  mutate(predicted=predict(fit), resid=y-predict(fit)) 
ggplot(fake2, aes(x=x, y=y, shape=as.factor(z), color=as.factor(z)) )+
  geom_point() +
  geom_line(aes(x=x, y=predicted, shape=as.factor(z), color=as.factor(z))) +
 xlab("Predicted Outcome (y-hat)") + ylab("Residual")

```
#### Create a Residual Plot 

In general it is good practice to plot the residuals vs. the fitted values. Below is all on one plot.

```{r }

#predicted is the y-hats, predicted values, using the model
#then we calculate the residuals as the actual value - the predicted value. 

ggplot(fake2, aes(x=predicted, y=resid, shape=as.factor(z), color=as.factor(z)) )+
  geom_point() +
  geom_hline(yintercept=0, linetype="dashed", color = "gray")
 
```
We also could make this graph as a facet graph.  Note that I am rewriting the book code using ggplot but I am fine with you coding your plots and things either way.  
```{r }

#predicted is the y-hats, predicted values, using the model
#then we calculate the residuals as the actual value - the predicted value. 

ggplot(fake2, aes(x=predicted, y=resid, shape=as.factor(z), color=as.factor(z)) )+
  geom_point() +
  geom_hline(yintercept=0, linetype="dashed", color = "gray") +
  facet_grid(. ~ as.factor(z), scales="free_x") +
 xlab("Predicted Outcome (y-hat)") + ylab("Residual")
 
```

Below is the book way.

```{r }
par(mfrow=c(1,2), mar=c(3,3,2,2), mgp=c(1.7,.5,0), tck=-.01)
for (i in 0:1){
  plot(range(x), range(y), type="n", xlab="Pre-treatment predictor, x", ylab="Outcome, y", main=paste("z =", i), bty="l")
  points(x[z==i], y[z==i], pch=20+i)
  abline(coef(fit)["(Intercept)"] + coef(fit)["z"]*i, coef(fit)["x"])
}
```
## More complicated model with multiple pre-treatment predictors.

#### Fake data

In this section we extend the model where we have K predictors.  The true model is:

$$ y = b_o + b_1x_1 + ... + b_kx_k + \theta z + error $$
Where z is binary.    Below is our code to generate fake data.

```{r }
N <- 100
K <- 10
X <- array(runif(N*K, 0, 1), c(N, K))
z <- sample(c(0, 1), N, replace=TRUE)
a <- 1
b <- 1:K
theta <- 10
sigma <- 5
y <- a + X %*% b + theta*z +  rnorm(N, 0, sigma)
fake <- data.frame(X=X, y=y, z=z)
```
Summarize our fake data to make sure it is what we expect.
```{r}
summary(fake)
#cor(fake)
```

#### Model
Then we fit the model and store the results in fit.
```{r }
fit <- stan_glm(y ~ X + z, data=fake, refresh = 0)
print(fit)
```

#### Plot Predictor vs Outcome

```{r }

#predicted is the y-hats, predicted values, using the model
#then we calculate the residuals as the actual value - the predicted value. 
fake2 <- fake %>%
  mutate(predicted=predict(fit), resid=y-predict(fit)) 
ggplot(fake2, aes(x=predicted, y=y, shape=as.factor(z), color=as.factor(z)) )+
  geom_point() +
  facet_grid(. ~ as.factor(z)) +
 xlab("Linear Predictor (y-hat)") + ylab("Outcome, y")
 
```
#### Book Way for Predictor vs. Outcome

```{r }
y_hat <- predict(fit)
par(mfrow=c(1,2), mar=c(3,3,2,2), mgp=c(1.7,.5,0), tck=-.01)
par(mfrow=c(1,2), pty="s")
for (i in 0:1){
  plot(range(y_hat,y), range(y_hat,y), type="n", xlab=expression(paste("Linear predictor, ", hat(y))), ylab="Outcome, y", main=paste("z =", i), bty="l")
  points(y_hat[z==i], y[z==i], pch=20+i)
abline(0, 1)
}
```


#### Plot Predictor vs Residual

```{r }

#predicted is the y-hats, predicted values, using the model
#then we calculate the residuals as the actual value - the predicted value. 

ggplot(fake2, aes(x=predicted, y=resid, shape=as.factor(z), color=as.factor(z)) )+
  geom_point() +
  geom_hline(yintercept=0, linetype="dashed", color = "gray") +
  facet_grid(. ~ as.factor(z), scales="free_x") +
 xlab("Predicted Outcome (y-hat)") + ylab("Residual")
 
```

#### What happens if there is Interaction but we miss it?
Below redo the simple example but add an interaction effect when you generate the fake data.
set the coefficient for the interaction to 20. What do you observe?




Once you fit the model with out an interaction, add the interaction and take a look at how the plots change.











#### What happens if there is non-constant variance?
Below redo the simple example but make the variance non-constant for different levels of x when you make the fake data.  The error term will be:

 error $\sim$ N(0, function(x){1+0.5*x}) 


